---
title: "R Logging Package Metrics"
author: "R. Mark Sharp"
date: "3/5/2021"
output:
  bookdown::pdf_document2: 
    includes:
        in_header: colortbl_package.sty
    toc: true
    toc_depth: 2
    number_sections: true
  bookdown::html_document2: 
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GeneKeepR Tutorial} 
  %\usepackage[UTF-8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown_notangle} 

---
  
```{r setup, echo = FALSE}
library(knitr)
library(ggplot2)
library(dlstats)
library(kableExtra)
library(rmsutilityr)
library(rvcheck)
library(stringi)
knitr::opts_chunk$set(echo = FALSE, cache = TRUE) # show all code

```
```{r get-stats}

pkgs <- c("futile.logger", "logging", "lgr", "log4r", "logger")
cran_stats_df <- cran_stats(pkgs, use_cache = TRUE)
licenses_df <- get_pkg_descriptions(pkgs, fields = "License")[["pkg_df"]]
bad_licenses <- c("AGPL")
require_approval_licenses <- c("LGPL")
cant_use <- licenses_df$Package[
  str_detect_any(licenses_df$License, pattern = bad_licenses, type = "fixed")]
require_approval <- licenses_df$Package[
  str_detect_any(licenses_df$License, pattern = require_approval_licenses, 
                 type = "fixed")]

get_vignette_counts <- function(pkgs) {
  vignettes <- vignette(package = pkgs)$results[ , 1]
  vignette_counts <- data.frame()
  for (pkg in pkgs) {
    vignette_counts <- 
      rbind(vignette_counts, 
            data.frame(Package = pkg,
                       Count = sum(vignettes == pkg)))
  }
  vignette_counts
} 
licenses_df <- get_pkg_descriptions(pkgs, fields = "License")$pkg_df

```


# Introduction
The goal of this document is to collect some readily available statistics of
the logging 
tools available in R to inform our decisions regarding which tool to use.
While `r numbers2words(length(cant_use))` of these `r length(pkgs)` packages 
(`r get_and_or_list(cant_use)`) can't be used by
some businesses because of current licensing 
(`r get_and_or_list(licenses_df$License[licenses_df$Package %in% cant_use])`),
`r ifelse(length(cant_use) > 1, "they", "it")` will not be excluded.

A more complete investigative approach has been presented by
[@https://doi.org/10.1002/ece3.5970]. This presentation was intended to use five 
of the criteria presented in that paper but not all have been done:

1   **Maturity** How long ago was its first release on CRAN   
      -  Not explicitely completed though the downloads plot
       (Figure \@ref(fig:get-downloads-plot)) indirectly provides that date.   
2   **Active development** Number of updates in past two years   
      -  Not completed.   
3   **Recently updated** Most recent update   
4   **Documentation available** Is there at least one vignette  
5   **License** Can it be used in a business setting without potential risk of 
    forcing disclosure of proprietary or sensitive code?   

The most important considerations in selecting a logging system is demonstrating 
it will work in your environment and it has the functionality you need.

# Most Recent Update

Once logging packages are recognized as feature complete, frequent updates are 
not expected.
However, updates may be forced by changes in to base `R` 
or their dependencies.
Note, based on the date of the most recent update,
`futile.logger` is apparently stable, resilient to changes 
in `R`, and was created with security features years before they were required 
by CRAN. It has very few dependencies (`utils`, `lambda.r`, and 
`futile.options`). Two of the three are by the same author and the
third is part of the core package set. See Table \@ref(tab:latest-updates).

```{r latest-updates}
suppressMessages(library(rvest))
library(ggplot2)

url = 'https://cran.r-project.org/web/packages/available_packages_by_date.html'

CRANpage <- read_html(url)
tbls <- html_nodes(CRANpage, "table") # since HTML is in table; no need to scrape td/tr elements
table_1 <- html_table(tbls[1], fill = TRUE)
loggers_pkgs <- table_1[[1]][table_1[[1]]$Package %in% pkgs, ]

#house cleaning
loggers_pkgs$Date <- as.Date(loggers_pkgs$Date)
caption <- "Most recent updates of packages."
kbl(loggers_pkgs[ , c("Package", "Date")],
    format = ifelse(knitr::is_latex_output(), "latex", "html"),
    booktabs = TRUE, caption = caption,
    row.names = FALSE,
    col.names = c("Package", "Latest Update"),
    longtable = TRUE) %>% 
  kable_styling(latex_options = c("repeat_header", "striped"), 
                font_size = ifelse(knitr::is_latex_output(), 8, 12)) %>%
  column_spec(1, width = "15em") %>%
  column_spec(2, width = "15em") 


```

# Downloads

The number of downloads a package experiences is an indicator of interest and
a weak surrogate of use. See Figure \@ref(fig:get-downloads-plot).

```{r get-downloads-plot, fig.cap = "Number of downloads per month of logging packages", cache = TRUE}
if (!is.null(cran_stats_df)) {
  ggplot(cran_stats_df, 
         aes(end, downloads, group = package, color = package)) +
    geom_line() + geom_point(aes(shape = package))
}

```

# Number of Reverse Dependencies

Reverse dependencies are a strong surrogate for use but not necessarily a 
strong surrogate for use in important packages or applications.
```{r define-count-dependencies, cache = TRUE}
count_dependencies <- 
  function(pkgs, 
           dependencies = "all",
           repos = "https://mirrors.nics.utk.edu/cran/") {
  dependencies_df <- data.frame()
  installed <- available.packages(repos = repos)
  for (i in seq_along(pkgs)) {
    dependent_pkgs <-
      tools::dependsOnPkgs(
        pkgs[i],
        dependencies = dependencies,
        installed = installed)
    dependencies_df <- rbind(dependencies_df, 
                       data.frame(pkg = pkgs[i],
                                  n_dependencies = length(dependent_pkgs)))
  }
  dependencies_df
}
```
Table \@ref(tab:reverse-dependencies-all) counts all dependences 
(`r get_and_or_list(c("Depends", "Imports", "LinkingTo", "Suggests", "Enhances"))`
), which is not very meaningful since these will include examples of the other
loggers in vignettes.

```{r reverse-dependencies-all, eval = TRUE, results = "asis", cache = TRUE}
dependencies <- c("Depends", "Imports", "LinkingTo", "Suggests", "Enhances")
dependency_counts <- count_dependencies(pkgs, dependencies = dependencies)
caption <- paste0("Logging packages and the number of packages currently on ",
                  "CRAN that are using each.")
kbl(dependency_counts,
    format = ifelse(knitr::is_latex_output(), "latex", "html"),
    booktabs = TRUE, caption = caption,
    row.names = FALSE,
    col.names = c("Package", "Number of Dependencies"),
    longtable = TRUE) %>% 
  kable_styling(latex_options = c("repeat_header", "striped"), 
                font_size = ifelse(knitr::is_latex_output(), 8, 12)) %>%
  column_spec(1, width = "15em") %>%
  column_spec(2, width = "15em") 

```

Table \@ref(tab:reverse-dependencies-required) shows only the counts of 
essential dependences 
(`r get_and_or_list(c("Depends", "Imports", "LinkingTo"))`), which is 
a more direct measure of use, but leaves out all uses that are not in CRAN.
The number of uses in CRAN is always going to be very small as logging is 
primarily used in applications and purpose driven scripts and in 
seldom desired in most general purpose packages.


```{r reverse-dependencies-required, eval = TRUE, results = "asis", cache = TRUE}
dependencies <- c("Depends", "Imports", "LinkingTo")
dependency_counts <- count_dependencies(pkgs, dependencies = dependencies)
caption <- paste0("Logging packages and the number of packages currently on ",
                  "CRAN that are using each.")
kbl(dependency_counts,
    format = ifelse(knitr::is_latex_output(), "latex", "html"),
    booktabs = TRUE, caption = caption,
    row.names = FALSE,
    col.names = c("Package", "Number of Dependencies"),
    longtable = TRUE) %>% 
  kable_styling(latex_options = c("repeat_header", "striped"), 
                font_size = ifelse(knitr::is_latex_output(), 8, 12)) %>%
  column_spec(1, width = "15em") %>%
  column_spec(2, width = "15em") 

```
# Availability of Vignettes
While not required, tutorial vignettes are a convenience. The number and even
the availability of tutorials within the parent package can be misleading since
the practice of including vignettes within packages was far less common in 
the past. Fortunately it has been growing in 
frequency. Counts of availabe vignettes are shown in Table 
\@ref(tab:vignette-counts).

```{r vignette-counts}
caption <- "Count of vignettes within the packages."
kbl(get_vignette_counts(pkgs),
    format = ifelse(knitr::is_latex_output(), "latex", "html"),
    booktabs = TRUE, caption = caption,
    row.names = FALSE,
    col.names = c("Package", "Number of Vignettes"),
    longtable = TRUE) %>% 
  kable_styling(latex_options = c("repeat_header", "striped"), 
                font_size = ifelse(knitr::is_latex_output(), 8, 12)) %>%
  column_spec(1, width = "15em") %>%
  column_spec(2, width = "15em") 

```


# Licensing

Table \@ref(tab:license) shows the licenses used by each of the `r length(pkgs)`
packages. Note that the LGPL licesnse is more permissive the GPL licenses.

```{r license}
caption <- "The license for each package"
kbl(licenses_df,
    format = ifelse(knitr::is_latex_output(), "latex", "html"),
    booktabs = TRUE, caption = caption,
    row.names = FALSE,
    col.names = c("Package", "License"),
    longtable = FALSE) %>% 
  kable_styling(latex_options = c("repeat_header", "striped"), 
                font_size = ifelse(knitr::is_latex_output(), 8, 12)) %>%
  column_spec(1, width = "15em") %>%
  column_spec(2, width = "15em")

```
